Fast Sorting on Modern Computers

### 2005-07-29 docs/tech_report/conclusions.tex
added some heading to conclusings
removed the word gonnet from code
finished up shellsort
added quicksort results
added comparisons with lamarca to the conclusions

\label{conclusions}
This section presents a summary of the results from previous chapters, lists
important results and discoveries, and suggests contributions made to computer
science by this project.

>>>>

\label{conclusions}
This section presents a summary of the results from previous chapters, compares
our results with LaMarca's, lists important results and discoveries, and
suggests contributions made to computer science by this research.

\section{Results summary}

\subsection{Order N squared sorts}
\subsection{Heapsorts}
\subsection{Mergesorts}
\subsection{Quicksorts}
\subsection{Radixsorts}
\subsection{Shellsorts}

\section{Comparison with LaMarca's results}

Over the course of this research. The vast majority of our results have
correlated exactly with LaMarca's. Many of our graphs are exactly the same
shape to LaMarca's, can be calculated exactly or alomst exactly using his
methods, or can be seen to be equivalent when scaled by an appropriate ratio
(due to different test parameters).

LaMarca's heapsort graphs were equivalent to ours in the areas of level 1
misses, instruction count and cycle count. Like LaMarca, we found that a four
heap lowered our instruction count. Our level 2 misses also concurred with
LaMarca, with only one difference: cache heapsort, being an out-of-place sort,
fills the cache fully with a smaller number of keys than base heapsort, which is
in place. LaMarca's results do not show this. Nor do they show a difference in
cache misses between these two sorts when the array does fit in the cache.
LaMarca's base heapsort have the characteristics of an out-of-place heapsort.
However, our results show that his implicit heap improvements have led to a
significantly faster heapsort.

LaMarca's initial improvements to base mergesort were due to reduce the
instruction count. We noted a reduction with each of the steps he prescribed. We
note too that each of his improvement reduce level 2 cache misses, and the shape
of our level 2 cache miss graphs exactly match his. We also note similar
characterists: LaMarca mentions a wobble in the instruction count graphs, due to
having to copy the final merge back; we note this same wobble.

However, we disagree with his results in several key places. Firstly, we note
that an inlined sorting method does not perform as fast as insertion sort. We
find it difficult to perform his improvements on either algorithm N or algorithm
S, and write our own mergesort instead. We expected that mergesort would have
half the instruction count of heapsortl we found instead they were very similar.

LaMarca did not include details of his tagging system in multi-mergesort. As a
result, we used a search, which greatly increased the instruction count. As a
result, mutli-mergesort did not perform as well as the mergesorts which had only
been tiled.

We note also that aligning the level 2 cache alone leads to worse
performance than not making any improvements at all, due to the thrashing of the
level 1 cache. When we aligned the arrays for the level 1 cache, the results
were much closer to LaMarca's results, and showed what an improvement his
techniques brought to those algorithms.

Our quicksort results, again, we very similar to LaMarca's. We showed an
increase is instruction count of 25\% due to multi-quicksort - very close to
LaMarca's 20\%.  The shape of our instruction count graphs were the same, and
switching to a binary search in multi-quicksort reduces the instruction count.
Like him, we found that the base quicksort had very few cache misses, and that
not using Sedgewick's insertion sort improvement reduces cache misses. We found
that multi-quicksort's k-way sort changes the slope of the graph, but our
result were not conclusive as to whether this would eventually make
multi-quicksort faster than the other versions. Our results do show it being
slower for all our test cases, due to the increase in instruction count.

LaMarca predicts two cache misses per key, which translated to one miss per key
with our 32 bit integers. We found this calculation to be true, as is his
calculation of a decrease of one cache miss in four: we note a reduction of one
cache miss in eight.

Where LaMarca found that using Sedgewicks insertion sort optimisation decreased
instruction count, we found the opposite. We also found that despite
improvements due to cache performance and instruction count, our base quicksort
performed faster than our tuned version. Finally, we found that although
changing multi-mergesort's sequential search to a binary search decreased the
instruction count, it greatly increased the number of branch mispredictions
incurred, and slowed the sort down accordingly.

LaMarca determines that radixsort could not be improved in the same was as the
comparison based sorts, and that due to poor cache performance, it ran more
slowly than quicksort. We showed, however, that one of the improvements applied
to mergesort could be applied to it, and that its resulting cache performance is
sufficiently good that its low instruction count and lack of branch
mispredictions make it the fastest sort in our survey.

\section{Important results and contributions}

%TODO lamarca does an overall comparison. we should do the same

\section{Instrucyion count results}

### 2005-07-29 docs/tech_report/heap.tex
added comparisons with lamarca to the conclusions

Slight differences exist though. The level 2 cache miss results here indicate
that while still sorting in the cache, the number of misses in the in-place
heapsort is less than that of the out-of-place sorts. LaMarca, however, has the
same amount of misses in both cases. Similarly, the results here indicate that
base heapsort experiences a sharp increase in cache misses. This increase comes
when the set size is twice that of cache mergesort when its sharp increase
occurs. LaMarca's results do not show this.

>>>>

Slight differences exist though. The level 2 cache miss results here indicate
that while still sorting in the cache, the number of misses in the in-place
heapsort is less than that of the out-of-place sorts. LaMarca, however, has the
same amount of misses in both cases. Similarly, the results here indicate that
base heapsort experiences a sharp increase in cache misses. Cache heaport also
has a similar increase, though LaMarca's results show that these occur at the
same position in the graph, in other words, when th set size is the same. Our
results, however, show that this occurs earlier in cache heapsort, due to the
fact that it is an out of place sort.
